\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-01-28 src/010chap1.tex
cleaned up tnotes

\subsection{Special case}
\label{sec:courseschedule}
As a special case of the study group accommodation problem, suppose
all the apartments are on the same street or if they are all lined up
on a single path, the street map becomes a tree that is just a
path. Then the problem becomes what is called an {\em interval
  assignment problem}. The idea of interval assignment may not be
obvious here; hence to see this, consider a different problem in
{\WSI} where the classes for these study groups courses need to be
scheduled during a day (or a week or any time period). Each study
group has a bunch of courses associated with it some of which may be
shared by two or more study groups. It is mandatory that a student who
is a member of a study group takes all the courses associated with
that group. There are slots during the day for classes to be held and
the problem is to allocate class slots to courses such that all the
classes of a study group are consecutive. It is debatable if this will
not hamper the attention span and memory retention rate of the
students but that is, regrettably, out of the scope of this
thesis. The parallels between this class allocation problem and the
accommodation problem can be seen as follows. The set $U$ here, are
the courses offered (say Course 101 {\coneohone}, Course 102
{\coneohtwo} and so on). In this variation of the problem, the
collection $\cF$ is the set of study groups but the study groups are
filled by course IDs (in place of students in the earlier
example). For instance, Course 101 is mandatory for all study groups
$\xLLL$, $\xGGG$, $\xBBB$, $\xTTT$ and Course 102 is mandatory for
only the $\xLLL$ group) and so on. The sequence of class slots for the
day (or week or any time period) is analogous to the street map in the
accommodation problem. It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree\tnote[minor]{Allowing any tree in this example could be seen as a
  scenario where there are parallel classes. A node falling in the
  path between two other nodes would mean that the corresponding is
  scheduled between the other two.}) is called an interval assignment
problem.

@@

\begin{definition}[\emph{\Cop}]
    \label{def:copmatrix}
  \begin{enumerate}
  \item \Cop
  \item Strong \cop
  \item \Cop order or \cop permutation
  \item \Cop for set systems\tnote{see mcc04}
  \end{enumerate}
\end{definition}

@@

\begin{definition}[\emph{\Crop}]
  \label{def:cropmatrix}
  \tnote[important]{define circ1 here. refer to it in chap 2 or
    duplicate it using slash def }
\end{definition}

@@

The construction of \PQtrees in \cite{bl76} draws on the close
relationship of matrices with \COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or
columns) of the matrix. This is constructed using an elaborate linear
time procedure and is also a test for planarity\tnote{check check
  check. both interval graph and planarity in this paper?}.  PQR trees
is a generalized data structure based on PQ trees \cite{mm96,mpt98}.
\cite{tm05} describes an improved algorithm to build PQR
trees. \tnote{improv in terms of what?}\cite{wlh02} describes the
simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive-ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great degree. \tnote{it
  reduces the solution search space. fill in the blanks.}

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed\tnote{in what way??} in \cite{co98}.  \tnote{expand
  on the COP order graph creation and it having to be bipartite for M
  to have COP. and thus an odd cycle being an evidence of no COP.}
\tnote{ where should this go?: (1) cite|jlm97 (application of PQ
  trees in graphics). (2) helly's theorem citation
  19XXdgk-Hellystheorem-Danzer-Gruenbaum-Klee}

@@

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of
sparse matrices where this number is much lesser than the number of
columns. $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but have at most $\Delta$ {\un}s in
any row. {\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with inducing COP on a matrix. In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP. Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\tnote{check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns?} but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem\tnote{elaborate - what are the results?}.

Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, \tnote{or is it a
  survey of another result?  check.} that finding a Tucker's forbidden
submatrix of at least $k$ rows is NP complete. \tnote{how is this
  different from booth's 75 result??}  \tnote{where should this
  go? cite|tz04 (approx submatrix with COP sparse matrices)}

\cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote{Theorem 2.1 in jkckv}
\tnote{(1) A connection of COP problem to the travelling salesman
  problem is also introduced. what does this mean? -- COP can be used
  as a tool to reorder $0.5T \le runs(M) \le T$. (2) The optimization
  version of the $k$-run problem, i.e. minimization of number of
  blocks of ones is proven to be NP complete by
  cite:k77}\tnote{are these two the same?} \tnote{what is the
  reduction?} \tnote{other problems similar to COP -- cite:ckl96
  (ILP, circ ones, one drop) -- cite:th98 (generalization of COP -
  minimax, biotonic column) Tucker}

@@

We see in Section~\ref{sec:motive} that pairwise intersection
cardinality preservation is necessary and sufficient for an interval
assignment to be feasible for a given hypergraph\footnote{A {\em
    hypergraph} is an alternate representation of a set system and
  will be used in this thesis.}~\footnote{See Section~\ref{ch:prelims} for the
  formal definition.} and thus is a characterization for COP
\cite{nsnrs09}. \tnote{Refer to the survey section that will list the
  theorems and/or lemma?} In our work we extend this characterization
and find that trio-wise intersection cardinality preservation makes a
tree path labeling\footnote{A {\em tree path labeling} $\cl$ is a
  bijection of paths from the target tree $T$ to the hyperedges in
  given hypergraph $\cF$.}~\footnotemark[4]  (TPL) feasible, which is a generalization of the COP
problem. This problem is defined as follows.

@@

\tnote{The following sentence seems out of place in this para} In
spite of this being a restricted case, we believe that our results are
of significant interest in understanding the nature of {\sc Graph
  Isomorphism} which is polynomial time solvable in interval graphs
while being hard on path graphs\cite{kklv10}. {\kstars} are a class
of trees which are in many ways very close to intervals or paths. Each
ray\footnote{The path from a leaf to the root, the vertex with highest
  degree, is called a {\em ray} of the \kstar.}~\footnotemark[4] are independent except for the
root\footnote{The vertex with maximum degree in a {\kstar} is called
  {\em root}.}~\footnotemark[4] and hence can be
considered as an independent interval till the root. Our algorithm
builds on this fact and uses the interval assignment
algorithm\cite{nsnrs09} up until ``reaching'' the root and then uses
the trio-wise intersection cardinality (the extra condition in ICPPL
that generalizes ICPIA) check to resolve the ambiguity about which ray
the algorithm should ``grow'' the solution into in the next iteration.

We also have an algorithm for solving {\CFTPL} with no restrictions on
the target tree or set size which runs in exponential time.  This
algorithm finds a path labeling from $T$ by decomposing the problem
into subproblems of finding path labeling of subsets of $\cF$ from
subtrees of $T$. Given the fact that binary matrices naturally
represent a set system (see Section~\ref{sec:motive}) and that the
{\em overlap}\tnote{Give informal definition of ``overlap''
  in footnote/endnote and link to sec:prelims. ADD the definition in prelim.}
relation between the sets involved is an obvious equivalence relation,
$\cF$ quite naturally partitions into equivalence classes known as
{\em overlap components}\tnote{Give informal definition of ``overlap
  components'' in footnote and link to sec:prelims. ADD the definition
  in prelim.}. In the context of COP, overlap components were used in
\cite{wlh02} and \cite{kklv10}. Moreover, \cite{nsnrs09} discovered
that these equivalence classes form a total order \tnote{Give informal
  definition of ``total order'' in footnote?}. We extend this to TPL
and find that when $\cF$ is a path hypergraph\footnote{If there exists
  an FTPL for a hypergraph $\cF$, it is called a path hypergraph.},
the classes can be partially ordered as an in-tree in polynomial
time. Once $\cF$ is ``broken'' into overlap components, one must
identify the subtree of $T$ that it needs to map to and this is the
hard part which is currently open to be solved in polynomial time.

\tnote{The connection of TPL to graph isomorphism will be made
  later in the document}

%\theendnotes

>>>>

\subsection{Special case}
\label{sec:courseschedule}
As a special case of the study group accommodation problem, suppose
all the apartments are on the same street or if they are all lined up
on a single path, the street map becomes a tree that is just a
path. Then the problem becomes what is called an {\em interval
  assignment problem}. The idea of interval assignment may not be
obvious here; hence to see this, consider a different problem in
{\WSI} where the classes for these study groups courses need to be
scheduled during a day (or a week or any time period). Each study
group has a bunch of courses associated with it some of which may be
shared by two or more study groups. It is mandatory that a student who
is a member of a study group takes all the courses associated with
that group. There are slots during the day for classes to be held and
the problem is to allocate class slots to courses such that all the
classes of a study group are consecutive. It is debatable if this will
not hamper the attention span and memory retention rate of the
students but that is, regrettably, out of the scope of this
thesis. The parallels between this class allocation problem and the
accommodation problem can be seen as follows. The set $U$ here, are
the courses offered (say Course 101 {\coneohone}, Course 102
{\coneohtwo} and so on). In this variation of the problem, the
collection $\cF$ is the set of study groups but the study groups are
filled by course IDs (in place of students in the earlier
example). For instance, Course 101 is mandatory for all study groups
$\xLLL$, $\xGGG$, $\xBBB$, $\xTTT$ and Course 102 is mandatory for
only the $\xLLL$ group) and so on. The sequence of class slots for the
day (or week or any time period) is analogous to the street map in the
accommodation problem. It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree) is called an interval assignment
problem.

@@

\begin{definition}[\emph{\Cop}]
    \label{def:copmatrix}
  \begin{enumerate}
  \item \Cop
  \item Strong \cop
  \item \Cop order or \cop permutation
  \end{enumerate}
\end{definition}

@@

\begin{definition}[\emph{\Crop}]
  \label{def:cropmatrix}
\end{definition}

@@

The construction of \PQtrees in \cite{bl76} draws on the close
relationship of matrices with \COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or
columns) of the matrix. This is constructed using an elaborate linear
time procedure and is also a test for planarity.  PQR trees is a
generalized data structure based on PQ trees \cite{mm96,mpt98}.
\cite{tm05} describes an improved algorithm to build PQR
trees. \cite{wlh02} describes the simpler algorithm for COT. Hsu also
invented PC trees \cite{wlh01} which is claimed to be much easier to
implement. \cite{nsnrs09} describes a characterization of
consecutive-ones property solely based on the cardinality properties
of the set representations of the columns (rows); every column (row)
is equivalent to a set that has the row (column) indices of the rows
(columns) that have one entries in this column (row). This is
interesting and relevant, especially to this thesis because it
simplifies COT to a great degree.

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed in \cite{co98}.

@@

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of
sparse matrices where this number is much lesser than the number of
columns. $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but have at most $\Delta$ {\un}s in
any row. {\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with inducing COP on a matrix. In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP. Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\temptext{ check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns? } but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem.

Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, \temptext{ or is it a
  survey of another result?  check.} that finding a Tucker's forbidden
submatrix of at least $k$ rows is NP complete. \temptext{ how is this
  different from booth's 75 result??}

\cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.

@@

We see in Section~\ref{sec:motive} that pairwise intersection
cardinality preservation is necessary and sufficient for an interval
assignment to be feasible for a given hypergraph\footnote{A {\em
    hypergraph} is an alternate representation of a set system and
  will be used in this thesis.}~\footnote{See Section~\ref{ch:prelims}
  for the formal definition.} and thus is a characterization for COP
\cite{nsnrs09}. In our work we extend this characterization and find
that trio-wise intersection cardinality preservation makes a tree path
labeling\footnote{A {\em tree path labeling} $\cl$ is a bijection of
  paths from the target tree $T$ to the hyperedges in given hypergraph
  $\cF$.}~\footnotemark[4] (TPL) feasible, which is a generalization
of the COP problem. This problem is defined as follows.

@@

\temptext{The following sentence seems out of place in this para} In
spite of this being a restricted case, we believe that our results are
of significant interest in understanding the nature of {\sc Graph
  Isomorphism} which is polynomial time solvable in interval graphs
while being hard on path graphs\cite{kklv10}. {\kstars} are a class
of trees which are in many ways very close to intervals or paths. Each
ray\footnote{The path from a leaf to the root, the vertex with highest
  degree, is called a {\em ray} of the \kstar.}~\footnotemark[4] are independent except for the
root\footnote{The vertex with maximum degree in a {\kstar} is called
  {\em root}.}~\footnotemark[4] and hence can be
considered as an independent interval till the root. Our algorithm
builds on this fact and uses the interval assignment
algorithm\cite{nsnrs09} up until ``reaching'' the root and then uses
the trio-wise intersection cardinality (the extra condition in ICPPL
that generalizes ICPIA) check to resolve the ambiguity about which ray
the algorithm should ``grow'' the solution into in the next iteration.

We also have an algorithm for solving {\CFTPL} with no restrictions on
the target tree or set size which runs in exponential time.  This
algorithm finds a path labeling from $T$ by decomposing the problem
into subproblems of finding path labeling of subsets of $\cF$ from
subtrees of $T$. Given the fact that binary matrices naturally
represent a set system (see Section~\ref{sec:motive}) and that the
{\em overlap} relation between the sets involved is an obvious
equivalence relation, $\cF$ quite naturally partitions into
equivalence classes known as {\em overlap components}. In the context
of COP, overlap components were used in \cite{wlh02} and
\cite{kklv10}. Moreover, \cite{nsnrs09} discovered that these
equivalence classes form a total order. We extend this to TPL and find
that when $\cF$ is a path hypergraph\footnote{If there exists an FTPL
  for a hypergraph $\cF$, it is called a path hypergraph.}, the
classes can be partially ordered as an in-tree in polynomial
time. Once $\cF$ is ``broken'' into overlap components, one must
identify the subtree of $T$ that it needs to map to and this is the
hard part which is currently open to be solved in polynomial time.
